Clone and graft:\\ Testing scientific applications as they are built

### 2014-12-03 turcksin-spring-mechanics/paper.tex
More.

First, floating point numbers do not enjoy the same properties as integers.

Talk about numdiff

>>>>

At the base of the problem is the fact that floating point numbers do not
enjoy the same mathematical properties as integers (or, in fact, as real
numbers). For example, if you take into account round-off, then $(a+b)+c$ is in
general \textit{not} the same as $a+(b+c)$ or $(a+c)+b$. You can try this out
with $a=1.0, b=-1.0, c=1e-20$: in double precision, you will get
$(a+b)+c=1e-20, a+(b+c)=(a+c)+b=0$. This matters because if you make changes
to the code base, upgrade the compiler, switch from one processor or machine
to another, or change compiler optimization level, floating point operations
will be executed in different order and the results will be different at the
level of roundoff. In other words, your tests will start to fail even though
nothing fundamentally changed -- but the output will be different.

Over the past 15 years, we have played with many ideas how to address this
problem. For example, we tried to just never output numbers to their full 16
digits of double precision. The idea is that if you only output 5 digits, then
roundoff surely can't play a role. The problem is that the numbers 1.000\,049
999\,999\,999\,9 and 1.000\,050\,000\,000\,000\,0 only differ by round-off,
but when rounded to 5 digits, you either get 1.000\,0 or 1.000\,1. One could
think that this is a contrived example, but it turns out that this happens
with rather frustrating regularity if you just output enough numbers -- as one
often does in scientific codes.

A second area that creates problems that codes often try to compute zeros, but
because of roundoff what is actually produced are numbers of the kind
4.863e-19. A human recognizes that this is a zero and, consequently, equal to
2.489e-19 if all of the numbers around it are on the order of one. But the
\texttt{diff} command will complain, and consequently signal a difference.

The solution to these problems that we have ultimately come to use
is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
\texttt{diff} but it recognizes if the text it compares in two files is
actually a number. If two numbers differ by less than some relative or
absolute tolerance, then they are considered equal even if the characters by
which they are represented in an output file are different. For example, you
would replace the \texttt{diff} command used previously by the
following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
10^{-6}$ or $|a-b|\le 10^{-8} \max\{|a|,|b|\}$:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
\end{lstlisting}
